---
title: "DV Assignment 2, Part 1: [Your Title Here]"
author: "Group 8: Alice Školoudíková, Robin Lux"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))

theme_set(theme_minimal(base_size = 11))
```

## Research question

**Do schools with similar test results give similar secondary-school advice, regardless of how disadvantaged their student population is?**

**Level: inference.**

**Target audience: secondary schools.**
Because our audience might not be familiar with the statistical concepts or even how does the collected data look like, the visualization should be simple and make the interpretation of the results very explicit. Also the visualization should have a clear message that is understandable both from the title and the plots themselves, for example by using a clear titles and axis labels and units that are familiar to our audience.

## Exploratory process

We operationalize the research question as: At the same share of pupils reaching the target levels in maths (1S) and reading (2F), do more disadvantaged schools (higher school weighting, *schoolweging*) give lower secondary-school advice?

Our exploration process was:

1. **Data preparation.** We combined the advice, reference-level and school-weighting files into one table with one row per school, and removed schools we could not use.
2. **Single variables.** We looked at the distribution of each variable to understand the data and spot problems, such as very small schools.
3. **Test results and advice.** We split schools into three equal groups by school weighting and plotted their advice against their test results. Because a school's advice can be summarised in several ways, we compared three summaries: the share of pupils advised HAVO or higher (and below HAVO), the mean advice on a scale from 1 (PRO) to 6 (VWO), and the share of pupils in each of six main tracks. We wanted to see whether the choice of summary changes the relationship between test results and advice.

### Data preparation

We kept regular primary schools only, joined the three files, and removed schools with more than one location (they cannot be matched to one school weighting), a missing school weighting, or no pupils in the reference levels or advice. DUO hides counts of 1–4 pupils as "<5"; we replaced these with 2.5 and will check later whether the results change when we use 1 or 4 instead. The table below shows how many schools remain after each step: we keep 5,457 of 6,385 schools (85%), and the largest loss comes from schools with more than one location.

```{r data-prep}

## Set value for "<5" to 2.5
hidden_value <- 2.5

## Create function that turns "<5" into 2.5 and text into numbers
to_number <- function(x) {
  x <- as.character(x)
  x[x == "<5"] <- hidden_value
  as.numeric(x)
}

## Order names of the ten advice tracks
track_cols <- c(
  "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT",
  "VMBO_GT", "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)
track_names <- c(
  "PRO", "VMBO-B", "VMBO-B/K", "VMBO-K", "VMBO-K/GT",
  "VMBO-GT", "VMBO-GT/HAVO", "HAVO", "HAVO/VWO", "VWO"
)

## Clean reference levels for maths and reading
levels_clean <- referentieniveaus |>
  # filter out special primary schools
  filter(SOORT_PO == "Bo") |>
  # turn pupil counts into numbers
  mutate(across(
    c(REKENEN_LAGER1F, REKENEN_1F, REKENEN_1S, LV_LAGER1F, LV_1F, LV_2F),
    to_number
  )) |>
  # count pupils and compute share at target level
  mutate(
    n_maths = REKENEN_LAGER1F + REKENEN_1F + REKENEN_1S,
    n_reading = LV_LAGER1F + LV_1F + LV_2F,
    pct_maths_1s = REKENEN_1S / n_maths,
    pct_reading_2f = LV_2F / n_reading
  ) |>
  # keep only the columns we need
  select(
    INSTELLINGSCODE, VESTIGINGSCODE, PROVINCIE, GEMEENTENAAM,
    n_maths, n_reading, pct_maths_1s, pct_reading_2f
  )

## Clean school advice
advice_clean <- schooladviezen |>
  # filter out special primary schools
  filter(SOORT_PO == "Bo") |>
  # turn pupil counts into numbers
  mutate(across(all_of(track_cols), to_number)) |>
  # count pupils, share of HAVO+ advice, mean advice (PRO = 1 ... VWO = 6)
  mutate(
    n_groep8 = PRO + VMBO_B + VMBO_B_K + VMBO_K + VMBO_K_GT +
      VMBO_GT + VMBO_GT_HAVO + HAVO + HAVO_VWO + VWO,
    pct_havo_plus = (HAVO + HAVO_VWO + VWO) / n_groep8,
    mean_advice = (
      1 * PRO + 2 * VMBO_B + 2.5 * VMBO_B_K + 3 * VMBO_K +
        3.5 * VMBO_K_GT + 4 * VMBO_GT + 4.5 * VMBO_GT_HAVO + 5 * HAVO +
        5.5 * HAVO_VWO + 6 * VWO
    ) / n_groep8
  ) |>
  # keep only the columns we need
  select(
    INSTELLINGSCODE, VESTIGINGSCODE, all_of(track_cols),
    n_groep8, pct_havo_plus, mean_advice
  )

## Clean school weighting
weighting_clean <- schoolweging |>
  # take school ID from OVT
  mutate(INSTELLINGSCODE = str_split_i(OVT, "\\|", 1)) |>
  # count locations per school
  add_count(INSTELLINGSCODE, name = "n_locations") |>
  # filter out schools with more than one location
  filter(n_locations == 1) |>
  # keep only the columns we need
  select(INSTELLINGSCODE, schoolweging, spreiding)

## Join advice with reference levels
step_levels <- advice_clean |>
  inner_join(levels_clean, by = c("INSTELLINGSCODE", "VESTIGINGSCODE"))

## Filter out schools with more than one location
step_single <- step_levels |>
  add_count(INSTELLINGSCODE, name = "n_locations") |>
  filter(n_locations == 1)

## Join with school weighting
step_weighting <- step_single |>
  inner_join(weighting_clean, by = "INSTELLINGSCODE")

## Filter out schools with missing school weighting
step_complete <- step_weighting |>
  filter(!is.na(schoolweging))

## Filter out schools with no pupils in reference levels or advice
schools <- step_complete |>
  filter(n_maths > 0, n_reading > 0, n_groep8 > 0)

## Create table of schools left after each step
sample_flow <- tibble(
  step = c(
    "All schools in advice file",
    "Regular schools only",
    "Matched to reference levels",
    "One location only",
    "Matched to school weighting",
    "School weighting not missing",
    "Pupils in reference levels and advice"
  ),
  schools = c(
    nrow(schooladviezen), nrow(advice_clean), nrow(step_levels),
    nrow(step_single), nrow(step_weighting), nrow(step_complete),
    nrow(schools)
  )
) |>
  mutate(lost = lag(schools) - schools)

knitr::kable(sample_flow)
```

### Histograms of variables

We first looked at the distribution of the variables we will work with. 

```{r exploratory-plot, fig.width = 12, fig.height = 8}

## Histogram of groep 8 size
p_size <- ggplot(schools, aes(x = n_groep8)) +
  geom_histogram(binwidth = 5, fill = "#1B3A5C", colour = "white") +
  labs(
    title = "Groep 8 size", x = "Number of groep 8 pupils",
    y = "Number of schools"
  )

## Histogram of school weighting
p_weighting <- ggplot(schools, aes(x = schoolweging)) +
  geom_histogram(binwidth = 1, fill = "#2C5F8A", colour = "white") +
  labs(
    title = "School weighting",
    x = "School weighting (higher = more disadvantaged)",
    y = "Number of schools"
  )

## Histogram of maths target level
p_maths <- ggplot(schools, aes(x = pct_maths_1s)) +
  geom_histogram(binwidth = 0.05, fill = "#3B7EA1", colour = "white") +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Maths target level",
    x = "Share of pupils at maths target level", y = "Number of schools"
  )

## Histogram of reading target level
p_reading <- ggplot(schools, aes(x = pct_reading_2f)) +
  geom_histogram(binwidth = 0.05, fill = "#4A90C2", colour = "white") +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Reading target level",
    x = "Share of pupils at reading target level", y = "Number of schools"
  )

## Histogram of HAVO+ advice
p_havo <- ggplot(schools, aes(x = pct_havo_plus)) +
  geom_histogram(binwidth = 0.05, fill = "#6BA3D0", colour = "white") +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "HAVO+ advice",
    x = "Share of pupils with HAVO+ advice", y = "Number of schools"
  )

## Find the highest bar of all five histograms
y_max <- max(
  layer_data(p_size)$count, layer_data(p_weighting)$count,
  layer_data(p_maths)$count, layer_data(p_reading)$count,
  layer_data(p_havo)$count
)

## Use the same y-axis (0 to highest bar) for all five histograms
same_y_axis <- scale_y_continuous(limits = c(0, y_max))
p_size <- p_size + same_y_axis
p_weighting <- p_weighting + same_y_axis
p_maths <- p_maths + same_y_axis
p_reading <- p_reading + same_y_axis
p_havo <- p_havo + same_y_axis

## Count pupils per advice track
advice_totals <- schools |>
  summarise(across(all_of(track_cols), sum)) |>
  pivot_longer(everything(), names_to = "track", values_to = "pupils") |>
  mutate(track = factor(track, levels = track_cols, labels = track_names))

## Bar chart of advice per track
p_tracks <- ggplot(advice_totals, aes(x = pupils, y = track)) +
  geom_col(fill = "#2F6F9F") +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Advice per track", x = "Number of pupils", y = NULL)

## Combine all six plots into one figure
plot_grid(p_size, p_weighting, p_maths, p_reading, p_havo, p_tracks, ncol = 3)
```

### Advice by test results and school disadvantage

To answer the research question, we plotted each school's advice against its test results, separately for maths and reading. Schools are split into three equal groups by school weighting, and each group has its own regression line.

```{r scatter-setup}
## Split schools into three equal groups by school weighting
schools <- schools |>
  mutate(
    disadvantage = ntile(schoolweging, 3),
    disadvantage = factor(
      disadvantage,
      levels = 1:3,
      labels = c("Least disadvantaged", "Middle", "Most disadvantaged")
    ),
    pct_below_havo = 1 - pct_havo_plus
  )

## Set colours for the three groups (colourblind-friendly)
disadv_colours <- c(
  "Least disadvantaged" = "#0072B2",
  "Middle" = "#009E73",
  "Most disadvantaged" = "#D55E00"
)

## Set x-axis titles for maths and reading
x_maths <- "Share of pupils at maths target level (1S)"
x_reading <- "Share of pupils at reading target level (2F)"

## Create function that makes one scatterplot
make_scatter <- function(data, x_col, y_col, title, x_label, y_label) {
  ggplot(
    data,
    aes(x = .data[[x_col]], y = .data[[y_col]], colour = disadvantage)
  ) +
    # one dot per school
    geom_point(alpha = 0.15, size = 1) +
    # one regression line per group
    geom_smooth(method = "lm", formula = y ~ x, linewidth = 1.2) +
    scale_colour_manual(values = disadv_colours) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    # show the full range (0-100%) on both axes
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(title = title, x = x_label, y = y_label) +
    theme_minimal(base_size = 16) +
    theme(legend.position = "none", plot.title = element_text(face = "bold"))
}

## Create function that makes two scatterplots (maths, reading) for one track
make_track_pair <- function(data, track_name) {
  # keep only this track
  track_data <- filter(data, track == track_name)
  y_label <- paste("Share of pupils with", track_name, "advice")
  list(
    make_scatter(
      track_data, "pct_maths_1s", "advice_share",
      title = paste(
        "Pupils with", track_name, "advice\nby school maths results"
      ),
      x_label = x_maths, y_label = y_label
    ),
    make_scatter(
      track_data, "pct_reading_2f", "advice_share",
      title = paste(
        "Pupils with", track_name, "advice\nby school reading results"
      ),
      x_label = x_reading, y_label = y_label
    )
  )
}

## Create one legend for all scatterplots (coloured square + text)
legend_data <- tibble(
  label = names(disadv_colours),
  x = c(0.30, 0.55, 0.68)
)
legend_row <- ggplot(legend_data) +
  # coloured squares
  geom_point(
    aes(x = x, y = 0), shape = 15, size = 9, colour = disadv_colours
  ) +
  # group names next to the squares
  geom_text(aes(x = x + 0.015, y = 0, label = label), hjust = 0, size = 6) +
  # legend title
  annotate(
    "text", x = 0.05, y = 0, label = "School disadvantage:",
    hjust = 0, size = 6
  ) +
  coord_cartesian(xlim = c(0, 1), ylim = c(-1, 1), expand = FALSE) +
  theme_void()
```

#### Split by HAVO+ and below HAVO advice

We first summarised the advice as the share of pupils with HAVO-or-higher advice, and the below HAVO.

```{r scatter-havo, fig.width = 14, fig.height = 13}
## Scatterplot of HAVO+ advice by maths target level
p_havo_maths <- make_scatter(
  schools, "pct_maths_1s", "pct_havo_plus",
  title = "Pupils with HAVO or higher advice\nby school maths results",
  x_label = x_maths, y_label = "Share of pupils with HAVO+ advice"
)

## Scatterplot of HAVO+ advice by reading target level
p_havo_reading <- make_scatter(
  schools, "pct_reading_2f", "pct_havo_plus",
  title = "Pupils with HAVO or higher advice\nby school reading results",
  x_label = x_reading, y_label = "Share of pupils with HAVO+ advice"
)

## Scatterplot of below HAVO advice by maths target level
p_below_maths <- make_scatter(
  schools, "pct_maths_1s", "pct_below_havo",
  title = "Pupils with advice below HAVO\nby school maths results",
  x_label = x_maths, y_label = "Share of pupils with below HAVO advice"
)

## Scatterplot of below HAVO advice by reading target level
p_below_reading <- make_scatter(
  schools, "pct_reading_2f", "pct_below_havo",
  title = "Pupils with advice below HAVO\nby school reading results",
  x_label = x_reading, y_label = "Share of pupils with below HAVO advice"
)

## Combine legend and the four plots (2 x 2)
havo_plots <- plot_grid(
  p_havo_maths, p_havo_reading, p_below_maths, p_below_reading,
  ncol = 2
)
plot_grid(legend_row, havo_plots, ncol = 1, rel_heights = c(0.05, 1))
```

#### Split by mean advice (1 = PRO to 6 = VWO) (to feedback)

Next, we summarised the advice as the mean advice per school, with tracks scored from 1 (PRO) to 6 (VWO) and split advice in between (e.g. HAVO/VWO = 5.5). We would like feedback on whether mean advice is a suitable summary, because it assumes equal distances between tracks which might not be appropriate.

```{r scatter-mean, fig.width = 14, fig.height = 7, message = FALSE}
## Set y-axis for mean advice (1 = PRO ... 6 = VWO, full range)
mean_y_axis <- list(
  scale_y_continuous(
    breaks = 1:6,
    labels = c("PRO", "VMBO-B", "VMBO-K", "VMBO-GT", "HAVO", "VWO")
  ),
  coord_cartesian(xlim = c(0, 1), ylim = c(1, 6))
)

## Scatterplot of mean advice by maths target level
p_mean_maths <- make_scatter(
  schools, "pct_maths_1s", "mean_advice",
  title = "Average advice per school\nby school maths results",
  x_label = x_maths, y_label = "Mean advice of the school"
) +
  mean_y_axis

## Scatterplot of mean advice by reading target level
p_mean_reading <- make_scatter(
  schools, "pct_reading_2f", "mean_advice",
  title = "Average advice per school\nby school reading results",
  x_label = x_reading, y_label = "Mean advice of the school"
) +
  mean_y_axis

## Combine legend and the two plots
mean_plots <- plot_grid(p_mean_maths, p_mean_reading, ncol = 2)
plot_grid(legend_row, mean_plots, ncol = 1, rel_heights = c(0.09, 1))
```

#### Split by six main tracks

Finally, we looked at the share of pupils in each of six main tracks (split advice counted as the lower track). At the same test results, it looks like more disadvantaged schools give more advice in the lower tracks (PRO to VMBO-K) and less VWO advice, especially among schools with high results. Visually, however, these figures are hard to read and navigate. We also tried to not use mean advice but all the 10 categories that are present in the data including all the split advices as separate categories but that made the figures simply worse without adding much extra information.

```{r scatter-six, fig.width = 14, fig.height = 31}
## Compute share per main track (split advice counts as the lower track)
six_long <- schools |>
  transmute(
    INSTELLINGSCODE, disadvantage, pct_maths_1s, pct_reading_2f,
    PRO = PRO / n_groep8,
    `VMBO-B` = (VMBO_B + VMBO_B_K) / n_groep8,
    `VMBO-K` = (VMBO_K + VMBO_K_GT) / n_groep8,
    `VMBO-GT` = (VMBO_GT + VMBO_GT_HAVO) / n_groep8,
    HAVO = (HAVO + HAVO_VWO) / n_groep8,
    VWO = VWO / n_groep8
  ) |>
  # one row per school and track
  pivot_longer(PRO:VWO, names_to = "track", values_to = "advice_share")

## Scatterplots of each main track (maths left, reading right)
six_plots <- c(
  make_track_pair(six_long, "PRO"),
  make_track_pair(six_long, "VMBO-B"),
  make_track_pair(six_long, "VMBO-K"),
  make_track_pair(six_long, "VMBO-GT"),
  make_track_pair(six_long, "HAVO"),
  make_track_pair(six_long, "VWO")
)

## Combine legend and the twelve plots (6 rows x 2 columns)
six_grid <- plot_grid(plotlist = six_plots, ncol = 2)
plot_grid(legend_row, six_grid, ncol = 1, rel_heights = c(0.02, 1))
```
