---
title: "DV Assignment 2: Same test results, different advice?"
author: "Group 8: Alice Školoudíková, Robin Lux"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

<!--
HOW TO KNIT THIS FILE

1. Open your group project in RStudio first (the `Project:` button in the
   top right should show your repo name). Keep this file in `report/`.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd in `report/`.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below sources scripts/00-packages.R and
   scripts/01-get-data.R, so packages and data are always loaded. The first
   knit downloads the data, so you need an internet connection.

If knitting fails:
- "could not find function": the package isn't loaded - add it to
  scripts/00-packages.R.
- "object not found": you created it in the Console, not in this file (or
  in a script that this file doesn't source).
- "cannot open file": check that the project is open, and use
  here("folder", "file") for paths instead of typing them out.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))
```

```{r data-prep, include = FALSE}
## Same data preparation as in Part 1 (see there for the sample flow)

## Set value for "<5" to 2.5
hidden_value <- 2.5

## Create function that turns "<5" into 2.5 and text into numbers
to_number <- function(x) {
  x <- as.character(x)
  x[x == "<5"] <- hidden_value
  as.numeric(x)
}

## Order names of the ten advice tracks
track_cols <- c(
  "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT",
  "VMBO_GT", "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)
track_names <- c(
  "PRO", "VMBO-B", "VMBO-B/K", "VMBO-K", "VMBO-K/GT",
  "VMBO-GT", "VMBO-GT/HAVO", "HAVO", "HAVO/VWO", "VWO"
)

## Clean reference levels for maths, reading and writing
levels_clean <- referentieniveaus |>
  # filter out special primary schools
  filter(SOORT_PO == "Bo") |>
  # turn pupil counts into numbers
  mutate(across(
    c(
      REKENEN_LAGER1F, REKENEN_1F, REKENEN_1S, LV_LAGER1F, LV_1F, LV_2F,
      TV_LAGER1F, TV_1F, TV_2F
    ),
    to_number
  )) |>
  # count pupils and compute share at target level per subject
  mutate(
    n_maths = REKENEN_LAGER1F + REKENEN_1F + REKENEN_1S,
    n_reading = LV_LAGER1F + LV_1F + LV_2F,
    n_writing = TV_LAGER1F + TV_1F + TV_2F,
    pct_maths_1s = REKENEN_1S / n_maths,
    pct_reading_2f = LV_2F / n_reading,
    pct_writing_2f = TV_2F / n_writing,
    # pooled index: share of all maths, reading and writing results that
    # reach the target level
    pct_target = (REKENEN_1S + LV_2F + TV_2F) /
      (n_maths + n_reading + n_writing)
  ) |>
  # keep only the columns we need
  select(
    INSTELLINGSCODE, VESTIGINGSCODE, PROVINCIE, GEMEENTENAAM,
    n_maths, n_reading, n_writing,
    pct_maths_1s, pct_reading_2f, pct_writing_2f, pct_target
  )

## Clean school advice
advice_clean <- schooladviezen |>
  # filter out special primary schools
  filter(SOORT_PO == "Bo") |>
  # turn pupil counts into numbers
  mutate(across(all_of(track_cols), to_number)) |>
  # count pupils, share of HAVO+ advice, mean advice (PRO = 1 ... VWO = 6)
  mutate(
    n_groep8 = PRO + VMBO_B + VMBO_B_K + VMBO_K + VMBO_K_GT +
      VMBO_GT + VMBO_GT_HAVO + HAVO + HAVO_VWO + VWO,
    pct_havo_plus = (HAVO + HAVO_VWO + VWO) / n_groep8,
    mean_advice = (
      1 * PRO + 2 * VMBO_B + 2.5 * VMBO_B_K + 3 * VMBO_K +
        3.5 * VMBO_K_GT + 4 * VMBO_GT + 4.5 * VMBO_GT_HAVO + 5 * HAVO +
        5.5 * HAVO_VWO + 6 * VWO
    ) / n_groep8
  ) |>
  # keep only the columns we need
  select(
    INSTELLINGSCODE, VESTIGINGSCODE, all_of(track_cols),
    n_groep8, pct_havo_plus, mean_advice
  )

## Clean school weighting
weighting_all <- schoolweging |>
  # drop rows without a school weighting (e.g. language classes, AZC
  # schools); they have no pupils and would otherwise count as an extra
  # location below
  filter(!is.na(schoolweging)) |>
  # take school ID from OVT
  mutate(INSTELLINGSCODE = str_split_i(OVT, "\\|", 1)) |>
  # count locations per school
  add_count(INSTELLINGSCODE, name = "n_locations")

## Keep schools with one location in the school weighting file
weighting_clean <- weighting_all |>
  filter(n_locations == 1) |>
  # keep only the columns we need
  select(INSTELLINGSCODE, schoolweging, spreiding)

## Join advice with reference levels
step_levels <- advice_clean |>
  inner_join(levels_clean, by = c("INSTELLINGSCODE", "VESTIGINGSCODE"))

## Filter out schools with more than one location
step_single <- step_levels |>
  add_count(INSTELLINGSCODE, name = "n_locations") |>
  filter(n_locations == 1)

## Keep schools that have a school weighting
step_found <- step_single |>
  semi_join(weighting_all, by = "INSTELLINGSCODE")

## Join with school weighting (one location in the weighting file only)
step_weighting <- step_found |>
  inner_join(weighting_clean, by = "INSTELLINGSCODE")

## Filter out schools with no pupils in reference levels or advice
schools <- step_weighting |>
  filter(n_maths > 0, n_reading > 0, n_writing > 0, n_groep8 > 0)

## Create table of schools left after each step
sample_flow <- tibble(
  step = c(
    "All schools in advice file",
    "Regular schools only",
    "Matched to reference levels",
    "One location in DUO files",
    "Has a school weighting",
    "One location in school weighting file",
    "Pupils in reference levels and advice"
  ),
  schools = c(
    nrow(schooladviezen), nrow(advice_clean), nrow(step_levels),
    nrow(step_single), nrow(step_found), nrow(step_weighting),
    nrow(schools)
  )
) |>
  mutate(lost = lag(schools) - schools)


## Split schools into three equal groups by school weighting
schools <- schools |>
  mutate(
    disadvantage = ntile(schoolweging, 3),
    disadvantage = factor(
      disadvantage,
      levels = 1:3,
      labels = c("Least disadvantaged", "Middle", "Most disadvantaged")
    )
  )

## Set colours for the three groups (colourblind-friendly)
disadv_colours <- c(
  "Least disadvantaged" = "#0072B2",
  "Middle" = "#009E73",
  "Most disadvantaged" = "#D55E00"
)

## Set x-axis title for the pooled test results
x_target <- "Share of results at target level (maths, reading and writing)"

## Create function that makes one scatterplot
make_scatter <- function(data, y_col, title, y_label, show_legend = FALSE) {
  ggplot(
    data,
    aes(x = .data$pct_target, y = .data[[y_col]], colour = .data$disadvantage)
  ) +
    # one dot per school
    geom_point(alpha = 0.15, size = 1) +
    # one regression line per group
    geom_smooth(method = "lm", formula = y ~ x, linewidth = 1.2) +
    scale_colour_manual(
      values = disadv_colours, name = "School disadvantage:"
    ) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    # show the full range (0-100%) on both axes
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(title = title, x = x_target, y = y_label) +
    # legend keys without the grey confidence band
    guides(colour = guide_legend(override.aes = list(fill = NA))) +
    theme_minimal(base_size = 16) +
    theme(
      legend.position = if (show_legend) "top" else "none",
      plot.title = element_text(face = "bold")
    )
}

## Create function that predicts y at one value of x (at) from the regression
## line of each group (by default the least and the most disadvantaged)
group_fit <- function(data, y_col, at,
                      groups = c("Least disadvantaged", "Most disadvantaged")) {
  fits <- vapply(groups, function(group) {
    x <- data$pct_target[data$disadvantage == group]
    y <- data[[y_col]][data$disadvantage == group]
    # least-squares line, same as geom_smooth(method = "lm")
    slope <- cov(x, y) / var(x)
    mean(y) + slope * (at - mean(x))
  }, numeric(1))
  fits
}

## Create function that highlights the gap between the least and the most
## disadvantaged schools at one test result with an arrow and a label
add_gap_label <- function(plot, fits, at, label_x, label_y, hjust) {
  label <- paste0(
    "At ", scales::percent(at), " of results at target level,\n",
    "the least disadvantaged schools advise ",
    scales::percent(fits[["Least disadvantaged"]], accuracy = 1), " HAVO+,\n",
    "the most disadvantaged schools ",
    scales::percent(fits[["Most disadvantaged"]], accuracy = 1)
  )
  plot +
    # arrow between the two regression lines
    annotate(
      "segment",
      x = at, xend = at,
      y = fits[["Most disadvantaged"]], yend = fits[["Least disadvantaged"]],
      linewidth = 1,
      arrow = arrow(ends = "both", length = unit(0.1, "inches"))
    ) +
    # explanation of the arrow
    annotate(
      "label",
      x = label_x, y = label_y, label = label, hjust = hjust, vjust = 0.5,
      size = 4.5, fill = "white"
    )
}
```

## Research question

*What are you investigating, and why does it matter? Who is the target audience for your final visualization, and how does that shape your choices? You can start from what you wrote in Part 1.*

## Exploratory process

*Describe how you explored the data before settling on your final visualization. Include at least one exploratory plot below (this can be the one from Part 1, or an updated one), and a short note on what it told you and how it influenced your final visualization.*

```{r exploratory-plot}
```

## Final visualization

*Your final, explanatory visualization goes below. Include a short write-up: what does it communicate, why is this the best way to communicate it, and how robust is the finding?*

```{r final-plot, fig.width = 16, fig.height = 15}
## ---- Top left: national averages -------------------------------------------

## National shares over all pupils of all regular primary schools (not only
## our analysis sample): HAVO+ advice and test results at target level
national_havo <- advice_clean |>
  summarise(share = sum(HAVO + HAVO_VWO + VWO) / sum(n_groep8)) |>
  pull(share)
national_target <- levels_clean |>
  summarise(
    share = sum(
      pct_maths_1s * n_maths + pct_reading_2f * n_reading +
        pct_writing_2f * n_writing,
      na.rm = TRUE
    ) / sum(n_maths + n_reading + n_writing)
  ) |>
  pull(share)

national <- tibble(
  measure = c(
    "Test results at\ntarget level",
    "Pupils with\nHAVO+ advice"
  ),
  share = c(national_target, national_havo)
) |>
  # first row on top
  mutate(measure = factor(measure, levels = rev(measure)))

## Bar chart: each bar on a grey 100% track, value printed next to it
p_national <- ggplot(national, aes(y = measure)) +
  geom_col(aes(x = 1), fill = "grey90", width = 0.6) +
  geom_col(aes(x = share), fill = "#1B3A5C", width = 0.6) +
  geom_text(
    aes(x = share, label = scales::percent(share, accuracy = 1)),
    hjust = 1.15, colour = "white", fontface = "bold", size = 7
  ) +
  scale_x_continuous(
    labels = scales::percent, limits = c(0, 1), expand = c(0, 0)
  ) +
  labs(
    title = "The Netherlands on average",
    subtitle = "All pupils of regular primary schools, 2024-2025",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 14, colour = "black")
  )

## ---- Top right: 100 pupils per group ---------------------------------------

## HAVO+ share each group's regression line predicts for a school whose test
## results equal the national average
group_havo <- group_fit(
  schools, "pct_havo_plus",
  at = national_target, groups = levels(schools$disadvantage)
)
group_people <- tibble(
  disadvantage = factor(names(group_havo), levels = names(group_havo)),
  havo = group_havo,
  n_havo = round(100 * group_havo)
)

## Outline of one pupil icon (head + shoulders) centred on (0, 0) in a
## 1 x 1 cell
circle <- seq(0, 2 * pi, length.out = 30)
shoulders <- seq(0, pi, length.out = 20)
pupil_head <- tibble(
  x = 0.15 * cos(circle), y = 0.24 + 0.15 * sin(circle), part = "head"
)
pupil_body <- tibble(
  x = c(0.3 * cos(shoulders), -0.3, 0.3),
  y = c(-0.12 + 0.28 * sin(shoulders), -0.42, -0.42),
  part = "body"
)
pupil_shape <- bind_rows(pupil_head, pupil_body)

## 100 pupils per group on a 10 x 10 grid (filled row by row from the top
## left); the first n_havo of them get HAVO+ advice
people <- group_people |>
  cross_join(tibble(pupil = 1:100)) |>
  mutate(
    col = (pupil - 1) %% 10,
    row = 9 - (pupil - 1) %/% 10,
    status = if_else(pupil <= n_havo, as.character(disadvantage), "other")
  ) |>
  cross_join(pupil_shape) |>
  mutate(x = col + x, y = row + y, id = paste(pupil, part))

## Percentage printed under each grid
people_labels <- group_people |>
  mutate(label = paste0(n_havo, " of 100 pupils\nget HAVO+ advice"))

p_people <- ggplot(people, aes(x = x, y = y)) +
  geom_polygon(aes(group = id, fill = status)) +
  geom_text(
    data = people_labels, aes(x = 4.5, y = -1.6, label = label),
    size = 5.5, lineheight = 0.9, fontface = "bold",
    colour = disadv_colours[as.character(people_labels$disadvantage)]
  ) +
  facet_wrap(~disadvantage) +
  scale_fill_manual(values = c(disadv_colours, other = "grey85")) +
  coord_fixed(ylim = c(-2.4, 9.5), clip = "off") +
  labs(
    title = paste0(
      "100 pupils at a school where ",
      scales::percent(national_target, accuracy = 1),
      " of test results reach the target level"
    ),
    subtitle = paste(
      "Same test results as the national average, but how many get",
      "HAVO+ advice\ndepends on how disadvantaged the school is"
    )
  ) +
  theme_void(base_size = 16) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(margin = margin(t = 4, b = 10)),
    strip.text = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    panel.spacing = unit(1.5, "lines")
  )

## ---- Bottom: HAVO+ advice by test results ----------------------------------

## Scatterplot from Part 1, with the gap between the least and the most
## disadvantaged schools marked at 60% of results at target level
gap_havo <- group_fit(schools, "pct_havo_plus", at = 0.6)
p_havo <- make_scatter(
  schools, "pct_havo_plus",
  title = "Pupils with HAVO or higher advice by school test results",
  y_label = "Share of pupils with HAVO+ advice", show_legend = TRUE
) |>
  add_gap_label(gap_havo, at = 0.6, label_x = 0.98, label_y = 0.1, hjust = 1)

## ---- Combine ---------------------------------------------------------------

top_row <- plot_grid(
  p_national, p_people,
  nrow = 1, rel_widths = c(1, 2)
)
plot_grid(
  top_row, p_havo,
  ncol = 1, rel_heights = c(1, 1.15)
)
```

## Conclusion

*What can you conclude from your visualization? What are the limitations of this analysis?*

---

**Before submitting:** run the lintr check on this file, and re-knit from a clean environment (`rm(list = ls())`, already done above) to make sure everything reproduces from scratch.
