---
title: "DV Assignment 2: Same test results, different advice?"
author: "Group 8: Alice Školoudíková, Robin Lux"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

<!--
HOW TO KNIT THIS FILE

1. Open your group project in RStudio first (the `Project:` button in the
   top right should show your repo name). Keep this file in `report/`.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd in `report/`.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below sources scripts/00-packages.R and
   scripts/01-get-data.R, so packages and data are always loaded. The first
   knit downloads the data, so you need an internet connection.

If knitting fails:
- "could not find function": the package isn't loaded - add it to
  scripts/00-packages.R.
- "object not found": you created it in the Console, not in this file (or
  in a script that this file doesn't source).
- "cannot open file": check that the project is open, and use
  here("folder", "file") for paths instead of typing them out.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))
```

```{r data-prep, include = FALSE}
## Same data preparation as in Part 1 (see there for the sample flow)

## Set value for "<5" to 2.5 (the midpoint of 1-4), unless the robustness
## check below reruns this chunk with another value
hidden_value <- getOption("hidden_value", default = 2.5)

## Create function that turns "<5" into hidden_value and text into numbers
to_number <- function(x) {
  x <- as.character(x)
  x[x == "<5"] <- hidden_value
  as.numeric(x)
}

## Order names of the ten advice tracks
track_cols <- c(
  "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT",
  "VMBO_GT", "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)
track_names <- c(
  "PRO", "VMBO-B", "VMBO-B/K", "VMBO-K", "VMBO-K/GT",
  "VMBO-GT", "VMBO-GT/HAVO", "HAVO", "HAVO/VWO", "VWO"
)

## Clean reference levels for maths, reading and writing
levels_clean <- referentieniveaus |>
  # filter out special primary schools
  filter(SOORT_PO == "Bo") |>
  # turn pupil counts into numbers
  mutate(across(
    c(
      REKENEN_LAGER1F, REKENEN_1F, REKENEN_1S, LV_LAGER1F, LV_1F, LV_2F,
      TV_LAGER1F, TV_1F, TV_2F
    ),
    to_number
  )) |>
  # count pupils and compute share at target level per subject
  mutate(
    n_maths = REKENEN_LAGER1F + REKENEN_1F + REKENEN_1S,
    n_reading = LV_LAGER1F + LV_1F + LV_2F,
    n_writing = TV_LAGER1F + TV_1F + TV_2F,
    pct_maths_1s = REKENEN_1S / n_maths,
    pct_reading_2f = LV_2F / n_reading,
    pct_writing_2f = TV_2F / n_writing,
    # pooled index: share of all maths, reading and writing results that
    # reach the target level
    pct_target = (REKENEN_1S + LV_2F + TV_2F) /
      (n_maths + n_reading + n_writing)
  ) |>
  # keep only the columns we need
  select(
    INSTELLINGSCODE, VESTIGINGSCODE, PROVINCIE, GEMEENTENAAM,
    n_maths, n_reading, n_writing,
    pct_maths_1s, pct_reading_2f, pct_writing_2f, pct_target
  )

## Clean school advice
advice_clean <- schooladviezen |>
  # filter out special primary schools
  filter(SOORT_PO == "Bo") |>
  # turn pupil counts into numbers
  mutate(across(all_of(track_cols), to_number)) |>
  # count pupils, share of HAVO+ advice, mean advice (PRO = 1 ... VWO = 6)
  mutate(
    n_groep8 = PRO + VMBO_B + VMBO_B_K + VMBO_K + VMBO_K_GT +
      VMBO_GT + VMBO_GT_HAVO + HAVO + HAVO_VWO + VWO,
    pct_havo_plus = (HAVO + HAVO_VWO + VWO) / n_groep8,
    mean_advice = (
      1 * PRO + 2 * VMBO_B + 2.5 * VMBO_B_K + 3 * VMBO_K +
        3.5 * VMBO_K_GT + 4 * VMBO_GT + 4.5 * VMBO_GT_HAVO + 5 * HAVO +
        5.5 * HAVO_VWO + 6 * VWO
    ) / n_groep8
  ) |>
  # keep only the columns we need
  select(
    INSTELLINGSCODE, VESTIGINGSCODE, all_of(track_cols),
    n_groep8, pct_havo_plus, mean_advice
  )

## Clean school weighting
weighting_all <- schoolweging |>
  # drop rows without a school weighting (e.g. language classes, AZC
  # schools); they have no pupils and would otherwise count as an extra
  # location below
  filter(!is.na(schoolweging)) |>
  # take school ID from OVT
  mutate(INSTELLINGSCODE = str_split_i(OVT, "\\|", 1)) |>
  # count locations per school
  add_count(INSTELLINGSCODE, name = "n_locations")

## Keep schools with one location in the school weighting file
weighting_clean <- weighting_all |>
  filter(n_locations == 1) |>
  # keep only the columns we need
  select(INSTELLINGSCODE, schoolweging, spreiding)

## Join advice with reference levels
step_levels <- advice_clean |>
  inner_join(levels_clean, by = c("INSTELLINGSCODE", "VESTIGINGSCODE"))

## Filter out schools with more than one location
step_single <- step_levels |>
  add_count(INSTELLINGSCODE, name = "n_locations") |>
  filter(n_locations == 1)

## Keep schools that have a school weighting
step_found <- step_single |>
  semi_join(weighting_all, by = "INSTELLINGSCODE")

## Join with school weighting (one location in the weighting file only)
step_weighting <- step_found |>
  inner_join(weighting_clean, by = "INSTELLINGSCODE")

## Filter out schools with no pupils in reference levels or advice
schools <- step_weighting |>
  filter(n_maths > 0, n_reading > 0, n_writing > 0, n_groep8 > 0)

## Create table of schools left after each step
sample_flow <- tibble(
  step = c(
    "All schools in advice file",
    "Regular schools only",
    "Matched to reference levels",
    "One location in DUO files",
    "Has a school weighting",
    "One location in school weighting file",
    "Pupils in reference levels and advice"
  ),
  schools = c(
    nrow(schooladviezen), nrow(advice_clean), nrow(step_levels),
    nrow(step_single), nrow(step_found), nrow(step_weighting),
    nrow(schools)
  )
) |>
  mutate(lost = lag(schools) - schools)

## Split schools into three equal groups by school weighting
schools <- schools |>
  mutate(
    disadvantage = ntile(schoolweging, 3),
    disadvantage = factor(
      disadvantage,
      levels = 1:3,
      labels = c("Least disadvantaged", "Middle", "Most disadvantaged")
    )
  )

## Set colours for the three groups (colourblind-friendly)
disadv_colours <- c(
  "Least disadvantaged" = "#0072B2",
  "Middle" = "#009E73",
  "Most disadvantaged" = "#D55E00"
)

## Set x-axis title for the pooled test results
x_target <- "Share of results at target level (maths, reading and writing)"

## Create function that makes one scatterplot
make_scatter <- function(data, y_col, title, y_label, show_legend = FALSE) {
  ggplot(
    data,
    aes(x = .data$pct_target, y = .data[[y_col]], colour = .data$disadvantage)
  ) +
    # one dot per school
    geom_point(alpha = 0.15, size = 1) +
    # one regression line per group
    geom_smooth(method = "lm", formula = y ~ x, linewidth = 1.2) +
    scale_colour_manual(
      values = disadv_colours, name = "School disadvantage:"
    ) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    # show the full range (0-100%) on both axes
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(title = title, x = x_target, y = y_label) +
    # legend keys without the grey confidence band
    guides(colour = guide_legend(override.aes = list(fill = NA))) +
    theme_minimal(base_size = 16) +
    theme(
      legend.position = if (show_legend) "top" else "none",
      plot.title = element_text(face = "bold")
    )
}

## Create function that predicts y at one value of x (at) from the regression
## line of each group (by default the least and the most disadvantaged),
## weighted by the number of groep 8 pupils
group_fit <- function(data, y_col, at,
                      groups = c("Least disadvantaged", "Most disadvantaged")) {
  fits <- vapply(groups, function(group) {
    in_group <- data$disadvantage == group
    group_data <- data.frame(
      x = data$pct_target[in_group],
      y = data[[y_col]][in_group],
      pupils = data$n_groep8[in_group]
    )
    # weighted least-squares line: schools with more pupils count more
    fit <- lm(y ~ x, data = group_data, weights = group_data$pupils)
    predict(fit, data.frame(x = at))
  }, numeric(1))
  fits
}

## Create function that highlights the gap between the least and the most
## disadvantaged schools at one test result with an arrow and a label
add_gap_label <- function(plot, fits, at, label_x, label_y, hjust) {
  label <- paste0(
    "At ", scales::percent(at), " of results at target level,\n",
    "the least disadvantaged schools advise ",
    scales::percent(fits[["Least disadvantaged"]], accuracy = 1), " HAVO+,\n",
    "the most disadvantaged schools ",
    scales::percent(fits[["Most disadvantaged"]], accuracy = 1)
  )
  plot +
    # arrow between the two regression lines
    annotate(
      "segment",
      x = at, xend = at,
      y = fits[["Most disadvantaged"]], yend = fits[["Least disadvantaged"]],
      linewidth = 1,
      arrow = arrow(ends = "both", length = unit(0.1, "inches"))
    ) +
    # explanation of the arrow
    annotate(
      "label",
      x = label_x, y = label_y, label = label, hjust = hjust, vjust = 0.5,
      size = 4.5, fill = "white"
    )
}
```

## Research question

**Do schools with similar test results give similar secondary-school advice, regardless of how disadvantaged their pupils are?** (RQ3, inference level)

Our audience is the people who give the advice: groep 8 teachers, especially at disadvantaged schools. The starting track matters, and switching later is hard. If two children with the same results get different advice because of their school, teachers should see that.

Teachers cannot be expected to know what "controlling for test results" means, so the figure makes that comparison for them. Disadvantaged schools have lower results on average, so a plain comparison would mislead. We compare schools at the same test results and count in pupils, because teachers think about the children in their class.

## Data and exploratory process

We kept regular primary schools, joined advice, reference levels and school weighting (*schoolweging*), and dropped schools with several locations (no single school weighting), leaving `r format(nrow(schools), big.mark = ",")` of `r format(nrow(advice_clean), big.mark = ",")` schools.

- **Test results:** one number per school, the share of all maths, reading and writing results at target level (1S for maths, 2F for reading and writing), as the question is about results as a whole.
- **Advice:** we tried the share of pupils with HAVO or higher advice (HAVO+) and a mean advice from 1 (PRO) to 6 (VWO). Both showed the same pattern, weaker for the mean, which also treats an ordinal scale as if the tracks were equally far apart. We kept HAVO+.
- **Hidden counts:** DUO shows counts of 1 to 4 pupils as "<5". We used 2.5 and checked 1 and 4 later.
- **Disadvantage:** three equal groups by school weighting.

```{r exploratory-plot, fig.width = 10, fig.height = 7.5}
## Exploratory plot (from Part 1): HAVO+ advice against test results, one
## regression line per third of the school weighting
explore_fits <- group_fit(schools, "pct_havo_plus", at = 0.6)
make_scatter(
  schools, "pct_havo_plus",
  title = "Exploratory: HAVO+ advice by test results and school disadvantage",
  y_label = "Share of pupils with HAVO+ advice", show_legend = TRUE
) |>
  add_gap_label(explore_fits, at = 0.6, label_x = 0.02, label_y = 0.9, hjust = 0)
```

What the plot showed, and what we changed:

1. The lines are clearly apart. The middle group sits in between and mostly adds clutter, so we compare only the most and the least disadvantaged third.
2. Reading a gap off two lines is the step we did not want to ask of teachers. The final figure picks one point, the national average, and shows what the lines predict there for 100 pupils.
3. The lines move apart as results improve, so one panel shows the gap over the whole range. Small schools are noisy (stripes at 50% and 100%), so we weight by groep 8 pupils.

## Final visualization

**This is our final, explanatory visualization.**

```{r final-plot, fig.width = 16, fig.height = 21, fig.showtext = TRUE}
## ---- Shared style ----------------------------------------------------------

## One font for the whole figure (downloaded from Google Fonts)
font_add_google("Inter", "inter")
showtext_auto()

## Colours: grey for context, one colour per highlighted group
ink <- "#1F1F1F"
muted <- "#4F4F4F"
dot_grey <- "#B8B8B8"
blue <- "#0072B2"
orange <- "#D55E00"
middle_grey <- "#8A8A8A"
gap_fill <- "#E9E9E9"

## One colour per disadvantage group (used in panels 1 and 3)
line_colours <- c(
  "Least disadvantaged" = blue,
  "Middle" = middle_grey,
  "Most disadvantaged" = orange
)

## Shared theme: white background, light horizontal gridlines only
theme_story <- function() {
  theme_minimal(base_size = 18, base_family = "inter") +
    theme(
      plot.title = element_text(
        face = "bold", size = 22, colour = ink, lineheight = 1.15,
        margin = margin(b = 10)
      ),
      plot.subtitle = element_text(
        size = 17, colour = muted, lineheight = 1.2, margin = margin(b = 24)
      ),
      axis.title = element_text(size = 17, colour = ink),
      axis.title.x = element_text(margin = margin(t = 16)),
      axis.title.y = element_text(margin = margin(r = 16)),
      axis.text = element_text(size = 15, colour = muted),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(colour = "#ECECEC", linewidth = 0.4),
      plot.margin = margin(20, 30, 50, 20)
    )
}

## Axes shared by panels 1 and 3
pct_axis <- function(name) {
  scale_x_continuous(
    name = name, labels = scales::percent, limits = c(0, 1),
    breaks = seq(0, 1, 0.2), expand = expansion(mult = 0.02),
    # keep arrows that end outside the plot
    oob = scales::oob_keep
  )
}
x_title <- "Test results (share of pupils reaching target level)"
y_title <- "Share of pupils receiving HAVO+ advice"

## ---- Numbers used in all three panels --------------------------------------

## National share of maths, reading and writing results at target level,
## over all pupils of all regular primary schools
national_target <- levels_clean |>
  summarise(
    share = sum(
      pct_maths_1s * n_maths + pct_reading_2f * n_reading +
        pct_writing_2f * n_writing,
      na.rm = TRUE
    ) / sum(n_maths + n_reading + n_writing)
  ) |>
  pull(share)
national_label <- scales::percent(national_target, accuracy = 1)

## HAVO+ share the regression line of the advantaged and the disadvantaged
## schools predicts at the national average (used in panels 1 and 2)
fits_avg <- group_fit(schools, "pct_havo_plus", at = national_target)

## ---- Panel 1: all schools ---------------------------------------------------

## Linear regression line over all schools, weighted by the number of groep 8
## pupils, over the middle 98% of the test results
all_fit <- lm(pct_havo_plus ~ pct_target, data = schools, weights = n_groep8)
all_line <- tibble(
  pct_target = quantile(schools$pct_target, c(0.01, 0.99))
) |>
  mutate(pct_havo_plus = predict(all_fit, newdata = pick(everything())))

## Correlation between test results and HAVO+ advice (weighted the same way),
## shown at the right end of the line
all_r <- cov.wt(
  select(schools, pct_target, pct_havo_plus),
  wt = schools$n_groep8, cor = TRUE
)$cor[1, 2]
line_label <- slice_max(all_line, pct_target, n = 1) |>
  mutate(label = paste("r =", format(round(all_r, 2), nsmall = 2)))

## Average share of pupils reaching target level of the advantaged and the
## disadvantaged schools (pooled over all their pupils, like the national
## average), to show that the two groups do not have the same results
group_avg <- schools |>
  filter(disadvantage != "Middle") |>
  group_by(disadvantage) |>
  summarise(
    x = sum(pct_target * (n_maths + n_reading + n_writing)) /
      sum(n_maths + n_reading + n_writing)
  ) |>
  mutate(
    label = paste(
      scales::percent(x, accuracy = 1),
      if_else(
        disadvantage == "Least disadvantaged", "advantaged", "disadvantaged"
      ),
      "\naverage"
    ),
    # label left of the lower average, right of the higher one
    hjust = if_else(disadvantage == "Least disadvantaged", 0, 1),
    nudge = if_else(disadvantage == "Least disadvantaged", 0.015, -0.015)
  )

## Where the arrows end, just right of panel 1 (next to the grids of panel 2)
arrow_end_x <- 1.53
arrow_end_y <- c("Least disadvantaged" = 0.69, "Most disadvantaged" = 0.24)

## Predicted HAVO+ share of the advantaged and the disadvantaged schools at
## the national average (the numbers shown in panel 2)
avg_points <- tibble(
  disadvantage = c("Least disadvantaged", "Most disadvantaged"),
  x = national_target,
  y = unname(fits_avg)
)

## Dot colour per school: blue for the least, orange for the most
## disadvantaged schools, grey for the middle group
dot_colours <- c(
  "Least disadvantaged" = blue,
  "Middle" = dot_grey,
  "Most disadvantaged" = orange
)

p1 <- ggplot(
  # middle group first, so the blue and orange dots are drawn on top
  arrange(schools, disadvantage != "Middle"),
  aes(x = pct_target, y = pct_havo_plus)
) +
  # one light dot per school, size = number of groep 8 pupils
  geom_point(aes(colour = disadvantage), size = 1.8, alpha = 0.12) +
  scale_colour_manual(values = dot_colours, guide = "none") +
  # national average on top of the dots, in dark ink so it stands out
  geom_vline(
    xintercept = national_target, linetype = "dashed", colour = ink,
    linewidth = 0.9
  ) +
  # average test results of the advantaged and the disadvantaged schools
  geom_vline(
    data = group_avg, aes(xintercept = x, colour = disadvantage),
    linetype = "dashed", linewidth = 0.9
  ) +
  # their labels on a white box, on two lines
  geom_label(
    data = group_avg,
    aes(x = x + nudge, y = 0.06, label = label, colour = disadvantage,
        hjust = hjust),
    size = 5.4, fontface = "bold", family = "inter", fill = "white",
    border.colour = "white", lineheight = 0.9
  ) +
  # regression line over all schools
  geom_line(
    data = all_line, aes(x = pct_target, y = pct_havo_plus),
    colour = ink, linewidth = 1.1
  ) +
  # arrows from the national average to the pupil grids in panel 2
  annotate(
    "curve",
    x = national_target + 0.01, y = fits_avg[["Least disadvantaged"]],
    xend = arrow_end_x, yend = arrow_end_y[["Least disadvantaged"]],
    curvature = -0.15, colour = ink, linewidth = 0.8,
    arrow = arrow(length = unit(0.15, "inches"), type = "closed")
  ) +
  annotate(
    "curve",
    x = national_target + 0.01, y = fits_avg[["Most disadvantaged"]],
    xend = arrow_end_x, yend = arrow_end_y[["Most disadvantaged"]],
    curvature = 0.25, colour = ink, linewidth = 0.8,
    arrow = arrow(length = unit(0.15, "inches"), type = "closed")
  ) +
  # points at the national average, where the arrows start
  geom_point(
    data = avg_points, aes(x = x, y = y, fill = disadvantage),
    shape = 21, size = 5, colour = "white", stroke = 1.5
  ) +
  # correlation at the end of the regression line
  geom_text(
    data = line_label,
    aes(x = pct_target + 0.015, y = pct_havo_plus, label = label),
    colour = ink, hjust = 0, size = 6, fontface = "bold", family = "inter"
  ) +
  scale_fill_manual(values = line_colours, guide = "none") +
  coord_cartesian(clip = "off") +
  # label of the national average on a white box, centred on top of its line
  annotate(
    "label",
    x = national_target, y = 1.08,
    label = paste(national_label, "national average"),
    hjust = 0.5, size = 5.4, fontface = "bold", colour = ink,
    family = "inter", fill = "white", label.size = 0
  ) +
  pct_axis(x_title) +
  scale_y_continuous(
    name = y_title, labels = scales::percent, limits = c(0, 1),
    # keep the national-average label above the plot
    oob = scales::oob_keep
  ) +
  labs(
    title = "1  Better test results, more HAVO+ advice",
    subtitle = "But schools with the same results still differ a lot"
  ) +
  theme_story() +
  # size legend above the plot
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.direction = "horizontal",
    legend.title = element_text(size = 15, colour = muted),
    legend.text = element_text(size = 14, colour = muted),
    # room under the legend for the national-average label
    legend.margin = margin(b = 38)
  )

## ---- Panel 2: 100 pupils at an average school ------------------------------

## One pupil grid each for the advantaged and the disadvantaged schools
grid_groups <- tibble(
  group = factor(
    c("More advantaged\nschools", "More disadvantaged\nschools"),
    levels = c("More advantaged\nschools", "More disadvantaged\nschools")
  ),
  colour = c(blue, orange),
  n_havo = round(100 * unname(fits_avg)),
  label = "pupils receive\nHAVO+ advice"
)

## Pupil icon (head + shoulders) in a 1 x 1 cell
circle <- seq(0, 2 * pi, length.out = 30)
shoulders <- seq(0, pi, length.out = 20)
pupil_shape <- bind_rows(
  tibble(x = 0.15 * cos(circle), y = 0.24 + 0.15 * sin(circle), part = "h"),
  tibble(
    x = c(0.3 * cos(shoulders), -0.3, 0.3),
    y = c(-0.12 + 0.28 * sin(shoulders), -0.42, -0.42),
    part = "b"
  )
)

## 100 pupils on a 10 x 10 grid, filled row by row from the bottom left
pupils <- grid_groups |>
  cross_join(tibble(pupil = 1:100)) |>
  mutate(
    col = (pupil - 1) %% 10,
    row = (pupil - 1) %/% 10,
    fill = if_else(pupil <= n_havo, colour, "#E4E4E4")
  ) |>
  cross_join(pupil_shape) |>
  mutate(x = col + x, y = row + y, id = paste(group, pupil, part))

p2 <- ggplot(pupils, aes(x = x, y = y)) +
  geom_polygon(aes(group = id, fill = fill)) +
  # big number and "pupils receive HAVO+ advice" to the right of each grid
  geom_text(
    data = grid_groups, aes(x = 10.5, y = 5.6, label = n_havo, colour = colour),
    hjust = 0, size = 15, fontface = "bold", family = "inter"
  ) +
  geom_text(
    data = grid_groups, aes(x = 10.6, y = 2.4, label = label),
    hjust = 0, size = 6.5, colour = muted, family = "inter", lineheight = 0.9
  ) +
  # grids under each other
  facet_wrap(~group, ncol = 1) +
  scale_fill_identity() +
  scale_colour_identity() +
  coord_fixed(xlim = c(-0.6, 16), ylim = c(-0.7, 9.7), clip = "off") +
  labs(
    title = "2  Same test results, different advice",
    subtitle = paste0(
      "Expected out of 100 pupils at national-average\nresults (",
      national_label, ")"
    )
  ) +
  theme_story() +
  theme(
    # larger subtitle than in the other panels
    plot.subtitle = element_text(
      size = 17, colour = muted, margin = margin(b = 24), lineheight = 1.2
    ),
    axis.text = element_blank(), axis.title = element_blank(),
    axis.title.x = element_blank(), axis.title.y = element_blank(),
    panel.grid.major.y = element_blank(), strip.text = element_blank(),
    panel.spacing = unit(5, "lines"),
    # extra space on the left to centre the grids in the right column
    plot.margin = margin(20, 30, 50, 135)
  )

## ---- Panel 3: the gap grows with test results ------------------------------

## Regression line (with 95% CI) for the advantaged and the disadvantaged
## schools, weighted by the number of groep 8 pupils, over the middle 98% of
## each group's test results
group_lines <- schools |>
  filter(disadvantage != "Middle") |>
  group_by(disadvantage) |>
  group_modify(function(d, key) {
    fit <- lm(pct_havo_plus ~ pct_target, data = d, weights = d$n_groep8)
    grid <- tibble(pct_target = seq(
      quantile(d$pct_target, 0.01), quantile(d$pct_target, 0.99),
      length.out = 50
    ))
    pred <- predict(fit, grid, interval = "confidence")
    bind_cols(grid, as_tibble(pred))
  }) |>
  ungroup()

## Shaded gap between the least and the most disadvantaged schools, where
## both groups have schools
gap_range <- group_lines |>
  group_by(disadvantage) |>
  summarise(lo = min(pct_target), hi = max(pct_target))
gap_x <- seq(max(gap_range$lo), min(gap_range$hi), length.out = 50)
gap_band <- tibble(pct_target = gap_x) |>
  mutate(
    top = map_dbl(pct_target, function(x) {
      group_fit(schools, "pct_havo_plus", at = x)[["Least disadvantaged"]]
    }),
    bottom = map_dbl(pct_target, function(x) {
      group_fit(schools, "pct_havo_plus", at = x)[["Most disadvantaged"]]
    })
  )

## Gap bars and labels (relative gap) at 40%, 60% and 80% of pupils reaching
## target level
gap_points <- tibble(pct_target = c(0.4, 0.6, 0.8)) |>
  mutate(
    top = map_dbl(pct_target, function(x) {
      group_fit(schools, "pct_havo_plus", at = x)[["Least disadvantaged"]]
    }),
    bottom = map_dbl(pct_target, function(x) {
      group_fit(schools, "pct_havo_plus", at = x)[["Most disadvantaged"]]
    }),
    # relative gap: how much more HAVO+ advice the advantaged schools give
    label = paste0(round(100 * (top - bottom) / bottom), "% more")
  )

## Line labels at the right end of each line
end_labels <- group_lines |>
  group_by(disadvantage) |>
  slice_max(pct_target, n = 1) |>
  ungroup() |>
  mutate(label = c("Advantaged schools", "Disadvantaged schools"))

## Smoothed distribution of the schools' test results per group, drawn at the
## bottom of panel 3 (same scale for both groups, highest peak = 14% of the
## y-axis)
density_height <- 0.14
school_density <- schools |>
  filter(disadvantage != "Middle") |>
  group_by(disadvantage) |>
  group_modify(function(d, key) {
    dens <- density(d$pct_target, from = 0, to = 1, n = 200)
    tibble(pct_target = dens$x, density = dens$y)
  }) |>
  ungroup() |>
  mutate(density_y = density / max(density) * density_height)

## Statistical test: does the slope differ between the advantaged and the
## disadvantaged schools? Weighted regression with an interaction term
outer_schools <- schools |>
  filter(disadvantage != "Middle") |>
  droplevels()
interaction_fit <- lm(
  pct_havo_plus ~ pct_target * disadvantage,
  data = outer_schools, weights = n_groep8
)
interaction_coef <- summary(interaction_fit)$coefficients[
  "pct_target:disadvantageMost disadvantaged",
]
interaction_p <- if (interaction_coef[["Pr(>|t|)"]] < 0.001) {
  "p < .001"
} else {
  paste("p =", round(interaction_coef[["Pr(>|t|)"]], 3))
}

p3 <- ggplot(group_lines, aes(x = pct_target)) +
  # distribution of the schools along the test results (grey area, coloured
  # outline per group)
  geom_ribbon(
    data = school_density,
    aes(ymin = 0, ymax = density_y, group = disadvantage),
    fill = dot_grey, alpha = 0.25
  ) +
  geom_line(
    data = school_density, aes(y = density_y, colour = disadvantage),
    linewidth = 0.8
  ) +
  geom_ribbon(
    data = gap_band, aes(ymin = bottom, ymax = top), fill = gap_fill
  ) +
  geom_ribbon(
    aes(ymin = lwr, ymax = upr, fill = disadvantage), alpha = 0.25
  ) +
  geom_line(aes(y = fit, colour = disadvantage), linewidth = 1.8) +
  # gap bar between the advantaged and the disadvantaged line
  geom_segment(
    data = gap_points, aes(xend = pct_target, y = bottom, yend = top),
    colour = ink, linewidth = 0.9,
    arrow = arrow(angle = 90, ends = "both", length = unit(0.06, "inches"))
  ) +
  geom_label(
    data = gap_points, aes(y = bottom - 0.04, label = label),
    size = 6.5, fontface = "bold", colour = ink, family = "inter",
    vjust = 1, fill = "white", label.size = 0,
    label.padding = unit(0.12, "lines"), lineheight = 0.9
  ) +
  geom_text(
    data = end_labels,
    aes(y = fit, label = label, colour = disadvantage),
    hjust = 0, nudge_x = 0.012, size = 6.5, fontface = "bold",
    family = "inter"
  ) +
  scale_colour_manual(values = line_colours, guide = "none") +
  scale_fill_manual(values = line_colours, guide = "none") +
  pct_axis(x_title) +
  scale_y_continuous(
    name = y_title, labels = scales::percent, limits = c(0, 1)
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "3  The advice gap grows as test results improve",
    subtitle = paste(
      "Expected share of HAVO+ advice for advantaged and",
      "disadvantaged schools"
    )
  ) +
  # explain the curves at the bottom
  annotate(
    "text", x = 1, y = 0.17, hjust = 1, colour = muted, size = 5.5,
    family = "inter", label = "How the schools' test results are spread"
  ) +
  theme_story() +
  theme(
    plot.margin = margin(30, 230, 10, 10),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

## ---- Combine ---------------------------------------------------------------

headline <- paste0(
  "<span style='color:", blue, "'>Advantaged schools</span> in the ",
  "Netherlands give more HAVO+ advice than<br>the <span style='color:",
  orange, "'>disadvantaged schools</span> give, even when students reach the ",
  "same test results."
)
caption <- paste0(
  "Source: DUO doorstroomtoets 2024-2025, ",
  format(nrow(schools), big.mark = ","), " regular primary schools. ",
  "HAVO+ includes HAVO, HAVO/VWO and VWO advice.\n",
  "School weighting (schoolweging) = CBS index of pupils' expected ",
  "educational disadvantage. Advantaged schools = the third with the lowest ",
  "school weighting,\ndisadvantaged schools = the third with the highest. ",
  "The middle third is not shown. Each dot is one school.\n",
  "Group 8: Alice Školouďíková, Robin Lux"
)

## Panels 1 and 2 next to each other, panel 3 under them at full width
top_row <- (p1 | p2) + plot_layout(widths = c(1.45, 1))
(top_row / p3) +
  plot_layout(heights = c(1, 1.25)) +
  plot_annotation(
    title = headline,
    caption = caption,
    theme = theme(
      plot.title = ggtext::element_markdown(
        family = "inter", face = "bold", size = 26, colour = ink,
        margin = margin(b = 44), lineheight = 1.2
      ),
      plot.caption = element_text(
        family = "inter", size = 14, colour = muted, hjust = 0,
        lineheight = 1.3, margin = margin(t = 30)
      ),
      plot.margin = margin(40, 32, 24, 32)
    )
  )
```

```{r report-numbers, include = FALSE}
## Numbers quoted in the text below the final figure
pct <- function(x) scales::percent(x, accuracy = 1)
avg_adv <- pct(group_avg$x[group_avg$disadvantage == "Least disadvantaged"])
avg_dis <- pct(group_avg$x[group_avg$disadvantage == "Most disadvantaged"])
n_adv <- round(100 * fits_avg[["Least disadvantaged"]])
n_dis <- round(100 * fits_avg[["Most disadvantaged"]])
gap_per_10 <- round(-10 * interaction_coef[["Estimate"]], 1)
n_multi <- sum(sample_flow$lost[c(4, 6)])
```

The headline states the conclusion; its coloured group names are the legend.

1. Better test results go with more HAVO+ advice (r = `r round(all_r, 2)`), but the groups start from different points: `r avg_adv` of results reach the target level at advantaged schools, `r avg_dis` at disadvantaged ones.
2. At the national average (`r national_label`), the lines predict `r n_adv` out of 100 pupils with HAVO+ advice at an advantaged school and `r n_dis` at a disadvantaged one.
3. The gap grows: advantaged schools give `r gap_points$label[1]` HAVO+ advice at 40% of results at target level and `r gap_points$label[3]` at 80%. The curves at the bottom show where schools are: disadvantaged schools with very high results are rare, so the end of the orange line rests on few schools.

Everything except the message is grey. Blue and orange are colourblind-safe and mean the same groups everywhere, the national average is the only reference point, and confidence bands show how certain the lines are.

**Statistical test.** A linear regression of the HAVO+ share on test results, group and their interaction, weighted by groep 8 pupils, gives an interaction of `r sprintf("%.2f", interaction_coef[["Estimate"]])` (`r interaction_p`): per 10 percentage points better results, the gap grows by about `r gap_per_10` percentage points.

### Robustness

```{r robustness}
## Fit the interaction model on the least and the most disadvantaged third and
## return the extra gap per 10 points of test results and the predicted HAVO+
## advice of both groups at the national average
robust_fit <- function(data, check, x_col = "pct_target", weighted = TRUE) {
  outer <- droplevels(data[data$disadvantage != "Middle", ])
  pupil_weights <- if (weighted) outer$n_groep8 else rep(1, nrow(outer))
  fit <- lm(
    reformulate(paste(x_col, "* disadvantage"), response = "pct_havo_plus"),
    data = outer, weights = pupil_weights
  )
  coefs <- summary(fit)$coefficients[
    paste0(x_col, ":disadvantageMost disadvantaged"),
  ]
  at_avg <- data.frame(
    disadvantage = c("Least disadvantaged", "Most disadvantaged")
  )
  at_avg[[x_col]] <- national_target
  pred <- unname(round(100 * predict(fit, at_avg)))
  data.frame(
    check = check,
    gap_growth = -10 * coefs[["Estimate"]],
    p = coefs[["Pr(>|t|)"]],
    advantaged = pred[1],
    disadvantaged = pred[2]
  )
}

## Rerun the data-prep chunk above with another value for "<5" and return
## its school table
schools_with <- function(value) {
  old <- options(hidden_value = value)
  on.exit(options(old))
  prep_env <- new.env()
  eval(parse(text = knitr::knit_code$get("data-prep")), envir = prep_env)
  prep_env$schools
}

## HAVO+ share with VMBO-GT/HAVO counted as HAVO+
schools_gt <- schools |>
  mutate(pct_havo_plus = (VMBO_GT_HAVO + HAVO + HAVO_VWO + VWO) / n_groep8)

robustness <- bind_rows(
  robust_fit(schools, "Main analysis (\"<5\" = 2.5)"),
  robust_fit(schools_with(1), "\"<5\" = 1"),
  robust_fit(schools_with(4), "\"<5\" = 4"),
  robust_fit(schools, "Unweighted", weighted = FALSE),
  robust_fit(filter(schools, n_groep8 >= 20), "Only schools with 20+ pupils"),
  robust_fit(schools_gt, "VMBO-GT/HAVO counted as HAVO+")
)

## Show the checks as a table
robustness |>
  transmute(
    Check = check,
    `Extra gap per 10 points` = paste(round(gap_growth, 1), "pp"),
    p = scales::pvalue(p),
    `HAVO+ out of 100 at national average` = paste(
      advantaged, "vs.", disadvantaged
    )
  ) |>
  knitr::kable(align = "lccc")

## Same model for each subject separately
subject_check <- bind_rows(
  robust_fit(schools, "Maths", x_col = "pct_maths_1s"),
  robust_fit(schools, "Reading", x_col = "pct_reading_2f"),
  robust_fit(schools, "Writing", x_col = "pct_writing_2f")
)
```

Direction and significance never change, only size. With "<5" = 1 the gap at the national average is `r robustness$advantaged[2] - robustness$disadvantaged[2]` pupils out of 100, with "<5" = 4 it is `r robustness$advantaged[3] - robustness$disadvantaged[3]`. Counting VMBO-GT/HAVO as HAVO+ slows the growth but keeps the gap. The gap also grows for each subject separately (maths `r round(subject_check$gap_growth[1], 1)`, reading `r round(subject_check$gap_growth[2], 1)`, writing `r round(subject_check$gap_growth[3], 1)` points per 10 points).

## Conclusion

No. At the national average, `r n_adv - n_dis` fewer pupils out of 100 get HAVO+ advice at a disadvantaged school, and the gap is largest where pupils do well. For teachers at disadvantaged schools: when your pupils score well, check whether your advice keeps up.

We compare schools, not pupils, so we cannot say which advice a particular child got. The figure shows a difference, not its cause: teachers use more than the test, and we cannot see that. We have one school year and left out `r n_multi` schools with several locations, so the pattern may differ for them or in other years. Within our data, it did not depend on any choice we tested.

## AI use

We used Claude Code (Claude Opus). It wrote most of the R code for the data preparation and the figure (styling, annotations, layout), rendered alternative versions of the figure for us to choose from, and helped draft parts of this report, which we rewrote and checked against our analysis.

We did not use it to decide the conclusion: we wrote the message and the headline ourselves, and kept our wording when it suggested changes. We gave it the research question, the codebook, the assignment and our rules (lintr style, the national average as the only reference point, schools rather than pupils).

We checked its work by knitting from a fresh clone and recomputing numbers in R. It did make mistakes: asked for a data preparation script, it first built a web page we had not asked for, and it kept suggesting extra curves and histograms that made the figure busier without strengthening the message. We rejected most of them.
